Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-15 paper_io_sim.tex
Edited methods

    We model page cache LRU lists as
    two lists of data blocks, an active list and an inactive list, both ordered by
    last access time (earliest first, see Figure~\ref{fig:lrulist}).
    As in the kernel, our simulator limits the size of the active list to
    twice the size of the inactive list, by moving least recently
    used data blocks from the active list to the inactive list~\cite{gorman2004understanding, linuxdev3rd2010}.

@@

            This prototype does not simulate  bandwidth sharing and thus does not support
            concurrency: it is limited to single-threaded applications running on systems
            with a single-core CPU. We used this prototype for a first validation of our simulation
            model against a real sequential application running on a real system.
            The source code this prototype is available at
            \url{https://github.com/big-data-lab-team/paper-io-simulation/tree/master/exp/pysim}. We used Python 3.7.

@@

        Our experiments compare real executions with our Python prototype,
        with the original \wrench, and with our \wrench-cache extension. We
        evaluate our page cache simulation in single-threaded and
        multi-threaded applications, accessing data on local and remote
        file systems. We use two applications: a synthetic one created to
        evaluate the simulation model, and a real one representative of
        neuroimaging data processing.

        The synthetic application, implemented in C, consists of three single-core,
        sequential tasks where each task reads the file produced by the
        previous task, increments every byte of this file to emulate real
        processing, and writes the resulting data to disk. Files are
        numbered by ascending creation times: File 1 is file read by Task 1, and File \texttt{i+1} is the file
        written by task \texttt{i}. In real executions, we clear the page
        cache before starting the application, to ensure comparable
        conditions among the runs. The anonymous memory used by application
        is released after each task, which we also simulate in the Python
        prototype and in WRENCH-cache.

        The real application is a workflow of the Nighres
        toolbox~\cite{huntenburg2018nighres} that implements skull
        stripping in high-resolution brain images, an operation common to
        many neuroimage analysis pipelines. The workflow is implemented as a Python
        script that calls Java image-processing routines. We used Python
        3.6, Java 8, and Nighres 1.3.0. We patched the
        skull stripping workflow to monitor task CPU times and inject them
        in the simulation, to focus our analysis on I/O times, the patched code is 
        available at \url{https://github.com/dohoangdzung/nighres}.

        The applications were executed on a dedicated cluster at
        Concordia University, with one login node, 9 compute nodes, and 4
        storage nodes connected with a network bandwidth of 25 Gbps. Each
        compute node has 2 $\times$ 16-core Intel(R) Xeon(R) Gold 6130 CPU
        @ 2.10GHz, 250~GiB of RAM, 6 $\times$ SSDs of 450~GiB each with the XFS
        file system, 378~GiB of tmpfs, and 126~GiB of devtmpfs file system.
        Nodes run CentOS~8.1 with NFS version 4. We use the \texttt{atop}
        and \texttt{collectl} tools to monitor and collect memory status
        and disk throughput.

        To parametrize our simulators, we benchmarked the
        memory, local disk, remote disk (NFS), and network bandwidths
        (Table~\ref{table:benchmark}). Since \simgrid, and thus \wrench, currently only supports
        symmetrical bandwidths, we use the mean of the read and write
        bandwidth values in our experiments.
            \begin{table}
            \centering
            \begin{tabularx}{\columnwidth}{ll
            >{\centering\arraybackslash}X
            >{\centering\arraybackslash}X
            >{\centering\arraybackslash}X}
            \toprule
                \multicolumn{2}{c}{Bandwidths}  & Cluster (real) & Python prototype & \wrench simulator\\
            \midrule
            \multirow{2}{*}{Memory}      & read  & 6860 & 4812 & 4812\\
                                         & write & 2764 & 4812 & 4812\\
            \multirow{2}{*}{Local disk}  & read  & 510  & 465  & 465\\
                                         & write & 420  & 465  & 465\\
            \multirow{2}{*}{Remote disk} & read  & 515  & -    & 445\\
                                         & write & 375  & -    & 445\\
            \multicolumn{2}{l}{Network}  & 3000  & -    & 3000\\
            \bottomrule
            \end{tabularx}
            \caption{Bandwidth benchmarks (MBps) and simulator configurations.
            The bandwidths used in the simulations are the average of the read and write bandwidths
            measured in the cluster.
            Remote disk and network accesses are not simulated in the Python simulator.}
            \label{table:benchmark}
            \end{table}

            As our focus is on I/O rather than compute, we measured
            application task CPU times on a cluster node
            (Table~\ref{table:cputime}), and used these durations in our
            simulations. For the Python simulator, we used CPU times
            directly in the simulation. For \wrench and \wrench-cache, we
            determined the corresponding number of flops on a 1~Gflops CPU
            and used these values in the simulation. The final simulated
            platform and application used in our experiments are available at
            commit \href{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}{ec6b43561b}.

@@

            Our first experiment (\textit{Exp single-threaded}) runs a
            single instance of the synthetic application on a single
            cluster node, with different input file sizes (20~GB, 50~GB,
            75~GB, 100~GB), and with all I/Os directed to the same local
            disk. The goal is to validate the simulation model in absence
            of any concurrent disk access.

            Our second experiment (\textit{Exp concurrent}) runs concurrent instances of the
            application on a single node, all application instances
            operating on different files of size 3~GB stored in the same
            local disk. We vary the number of
            concurrent application instances from 1 to 32 since cluster
            nodes have 32 CPU cores.

            Our third experiment (\textit{Exp NFS}) occurs in the same
            configuration as the previous one, albeit reading and writing
            on a 50-GiB NFS-mounted partition of a 450-GiB remote disk of
            another compute node. As is commonly configured in HPC
            environments to avoid data loss, there is no client write cache
            and the server cache is configured as writethrough instead of
            writeback. NFS client and server read caches are enabled. 
            Therefore, all the writes happen at disk bandwidth, but
            reads can benefit from cache hits.

            Our fourth and final experiment (\textit{Exp Nighres})
            simulates the Nighres CRUISE cortical reconstruction workflow, which consists 
            of four steps. The first two steps are skull stripping and MGDM segmentation, 
            which are in this Nighres \href{{https://nighres.readthedocs.io/en/latest/auto_examples/example_01_tissue_classification.html}}{tissue classification example}. 
            The last two tasks are region extraction and CRUISE cortical reconstruction 
            in the   
            \href{https://nighres.readthedocs.io/en/latest/auto_examples/example_02_cortical_depth_estimation.html}{cortical depth estimation example}.
            \tristan{describe data, and if the execution was single node or
            distributed.}

@@

    \section{Conclusion}
    \label{discussion}
    We designed a model of the Linux page cache and implemented it in the
    \simgrid-based \wrench library to simulate distributed applications.
    Evaluation results show that our model improves simulation accuracy
    substantially, decreasing absolute relative simulation errors up to
    9$\times$ (see results of the single-threaded experiment). The
    availability of asymmetrical disk bandwidths in the forthcoming
    \simgrid release will further improve the accuracy of our simulations.
    Our page cache model is publicly available in the \wrench GitHub
    repository.

    Page cache simulation would be instrumental in a number of studies. For
    instance, it is now common for HPC clusters to run applications in
    Linux control groups (cgroups), where resource consumption is limited,
    including memory and therefore page cache usage. Using our simulator,
    it would be possible to study the interaction between memory allocation
    and I/O performance, for instance to improve scheduling algorithms or
    avoid page cache starvation~\cite{zhuang2017}. Our simulator could also
    be utilized to evaluate solutions that reduce the impact of network
    file transfers on distributed applications, such as burst
    buffers~\cite{ferreiradasilva-fgcs-bb-2019}, hierarchical file
    systems~\cite{islam2015triple}, active storage~\cite{5496981}, or
    specific hardware architectures~\cite{hayot2020performance}.

    However, I/Os remain much more complex than is currently available in
    simulation models, which could drastically limit the accuracy of simulations.
    Relevant extensions to this work could include more
    accurate descriptions of anonymous memory usage by applications. 
    With the page cache, I/O times are indeed strongly
    dependent on the amount of available memory, which is itself impacted
    by the use of anonymous memory. File access patterns, 
    including random I/O, might also be worth including in application models,
    as they directly impact page cache content.
    % readahead and persistent storage could also be adeed
    % Simulation results could be made even more accurate by a deeper
    % investigation of page cache flushing and eviction order.

        \section{Acknowledgments}
The computing platform used in the experiments was obtained with funding
from the Canada Foundation for Innovation.

>>>>

    We model page cache LRU lists as
    two lists of data blocks, an active list and an inactive list, both ordered by
    last access time (earliest first, Figure~\ref{fig:lrulist}).
    As in the kernel, our simulator limits the size of the active list to
    twice the size of the inactive list, by moving least recently
    used data blocks from the active list to the inactive list~\cite{gorman2004understanding, linuxdev3rd2010}.

@@

            This prototype does not simulate  bandwidth sharing and thus does not support
            concurrency: it is limited to single-threaded applications running on systems
            with a single-core CPU. We used this prototype for a first validation of our simulation
            model against a real sequential application running on a real system.
            The Python 3.7 source code this prototype is available at
            \url{https://github.com/big-data-lab-team/paper-io-simulation/tree/master/exp/pysim}.

@@

        Our experiments compare real executions with our Python prototype,
        with the original \wrench, and with our \wrench-cache extension. We
        evaluate our page cache simulation in single-threaded and
        multi-threaded applications, accessing data on local and remote
        file systems. We use two applications: a synthetic one, designed to
        evaluate the simulation model, and a real one, representative of
        neuroimaging data processing.

        The synthetic application, implemented in C, consists of three single-core,
        sequential tasks where each task reads the file produced by the
        previous task, increments every byte of this file to emulate real
        processing, and writes the resulting data to disk. Files are
        numbered by ascending creation times: File 1 is the file read by Task 1, and file \texttt{i+1} is the file
        written by task \texttt{i}. In real executions, we clear the page
        cache before starting the application, to ensure comparable
        conditions among the runs. The anonymous memory used by application
        is released after each task, which we also simulate in the Python
        prototype and in WRENCH-cache.

        The real application is a workflow of the Nighres
        toolbox~\cite{huntenburg2018nighres} that implements
        cortical reconstruction from brain images in four steps: skull stripping, tissue classification, 
        region extraction, and cortical reconstruction. More information 
        on this applicatoin is available in the Nighres documentation at \url{https://nighres.readthedocs.io/en/latest/auto_examples/example_02_cortical_depth_estimation.html}.
        The workflow is implemented as a Python
        script that calls Java image-processing routines. We used Python
        3.6, Java 8, and Nighres 1.3.0. We patched the
        application to remove lazy data loading, which made CPU time difficult to separate from I/O time, 
        and to monitor task CPU times, to inject them in the simulators. The patched code is 
        available at \url{https://github.com/dohoangdzung/nighres}.

        The applications were executed on a dedicated cluster at
        Concordia University, with one login node, 9 compute nodes, and 4
        storage nodes connected with a network bandwidth of 25 Gbps. Each
        compute node has 2 $\times$ 16-core Intel(R) Xeon(R) Gold 6130 CPU
        @ 2.10GHz, 250~GiB of RAM, 6 $\times$ SSDs of 450~GiB each with the XFS
        file system, 378~GiB of tmpfs, and 126~GiB of devtmpfs file system.
        Nodes run CentOS~8.1 with NFS version 4. We used the \texttt{atop}
        and \texttt{collectl} tools to monitor and collect memory status
        and disk throughput.

        To parametrize the simulators, we benchmarked the
        memory, local disk, remote disk (NFS), and network bandwidths
        (Table~\ref{table:benchmark}). Since \simgrid, and thus \wrench, currently only supports
        symmetrical bandwidths, we use the mean of the read and write
        bandwidth values in our experiments.
            \begin{table}
            \centering
            \begin{tabularx}{\columnwidth}{ll
            >{\centering\arraybackslash}X
            >{\centering\arraybackslash}X
            >{\centering\arraybackslash}X}
            \toprule
                \multicolumn{2}{c}{Bandwidths}  & Cluster (real) & Python prototype & \wrench simulator\\
            \midrule
            \multirow{2}{*}{Memory}      & read  & 6860 & 4812 & 4812\\
                                         & write & 2764 & 4812 & 4812\\
            \multirow{2}{*}{Local disk}  & read  & 510  & 465  & 465\\
                                         & write & 420  & 465  & 465\\
            \multirow{2}{*}{Remote disk} & read  & 515  & -    & 445\\
                                         & write & 375  & -    & 445\\
            \multicolumn{2}{l}{Network}  & 3000  & -    & 3000\\
            \bottomrule
            \end{tabularx}
            \caption{Bandwidth benchmarks (MBps) and simulator configurations.
            The bandwidths used in the simulations are the average of the read and write bandwidths
            measured in the cluster.
            Remote disk and network accesses are not simulated in the Python simulator.}
            \label{table:benchmark}
            \end{table}

            As our focus is on I/O rather than compute, we measured
            application task CPU times on a cluster node
            (Table~\ref{table:cputime}), and used these durations in our
            simulations. For the Python prototype, we injected CPU times
            directly in the simulation. For \wrench and \wrench-cache, we
            determined the corresponding number of flops on a 1~Gflops CPU
            and used these values in the simulation. The final simulated
            platform and application used in our experiments are available at
            commit \href{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}{ec6b43561b}.

@@

            Our first experiment runs a
            single instance of the synthetic application on a single
            cluster node, with different input file sizes (20~GB, 50~GB,
            75~GB, 100~GB), and with all I/Os directed to the same local
            disk. The goal is to validate the simulation model in absence
            of any concurrent disk access.

            Our second experiment runs concurrent instances of the
            application on a single node, all application instances
            operating on different files of size 3~GB stored in the same
            local disk. We vary the number of
            concurrent application instances from 1 to 32 since cluster
            nodes have 32 CPU cores.

            Our third experiment occurs in the same
            configuration as the previous one, albeit reading and writing
            on a 50-GiB NFS-mounted partition of a 450-GiB remote disk of
            another compute node. As is commonly configured in HPC
            environments to avoid data loss, there is no client write cache
            and the server cache is configured as writethrough instead of
            writeback. NFS client and server read caches are enabled. 
            Therefore, all the writes happen at disk bandwidth, but
            reads can benefit from cache hits.

            Our fourth and final experiment runs the Nighres workflow on
            data from participant 0027430 in the dataset of the Max Planck Institute for
            Human Cognitive and Brain Sciences
            available at
            \url{http://dx.doi.org/10.15387/fcp_indi.corr.mpg1}. The data
            amounts to 2.3~GB and was downloaded from
            \url{http://datasets.datalad.org/?dir=/corr/RawDataBIDS/MPG_1/sub-0027430}.
            The workflow was run on a single node of the cluster, using a
            single local disk.

@@

    \section{Conclusion}
    \label{discussion}
    We designed a model of the Linux page cache and implemented it in the
    \simgrid-based \wrench library to simulate distributed applications.
    Evaluation results show that our model improves simulation accuracy
    substantially, reducing absolute relative simulation errors up to
    9$\times$ (see results of the single-threaded experiment). The
    availability of asymmetrical disk bandwidths in the forthcoming
    \simgrid release will further improve these results.
    Our page cache model is publicly available in the \wrench GitHub
    repository.

    Page cache simulation can be instrumental in a number of studies. For
    instance, it is now common for HPC clusters to run applications in
    Linux control groups (cgroups), where resource consumption is limited,
    including memory and therefore page cache usage. Using our simulator,
    it would be possible to study the interaction between memory allocation
    and I/O performance, for instance to improve scheduling algorithms or
    avoid page cache starvation~\cite{zhuang2017}. Our simulator could also
    be utilized to evaluate solutions that reduce the impact of network
    file transfers on distributed applications, such as burst
    buffers~\cite{ferreiradasilva-fgcs-bb-2019}, hierarchical file
    systems~\cite{islam2015triple}, active storage~\cite{5496981}, or
    specific hardware architectures~\cite{hayot2020performance}.

    However, I/Os remain much more complex than is currently available in
    simulation models, which could drastically limit the accuracy of simulations.
    Relevant extensions to this work could include more
    accurate descriptions of anonymous memory usage by applications. 
    With the page cache, I/O times are indeed strongly
    dependent on the amount of available memory, which is itself impacted
    by the use of anonymous memory. File access patterns, 
    including random I/O, might also be worth including in application models,
    as they directly affect page cache content.
    % readahead and persistent storage could also be adeed
    % Simulation results could be made even more accurate by a deeper
    % investigation of page cache flushing and eviction order.

        \section{Acknowledgments}
The computing platform used in the experiments was obtained with funding
from the Canada Foundation for Innovation. This work was partially supported
by NSF contracts \#1923539 and \#1923621.
